---
citecolor: Blue
classoption: table
colorlinks: yes
fontsize: 10pt
link-citations: yes
linkcolor: Blue
output:
  beamer_presentation:
    includes:
      after_body: doc_suffix.tex
      before_body: doc_prefix.tex
      in_header: header.tex
    slide_level: 3
urlcolor: Maroon
---

```{r options, include=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(echo=TRUE, cache=TRUE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```

<!-- ================================================================= -->
<!-- PART 1: Rcpp* packages -->
<!-- ================================================================= -->

# Short presentation of Rcpp* packages

## Rcpp: extending R with C++

### Rcpp R package

- **Rcpp** is an R package to extend R with C++ code
- Main advantage: C++ is fast, it accelerates R (see next sections)
- Written by **Dirk EDDELBUETTEL** and **Romain FRANCOIS**
- <http://www.rcpp.org/>

### Simple Rcpp example

**C++ code** (in file `Code/addition.cpp`)
```{Rcpp Rcpp-addition, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
int addition(int a, int b) {
  return a + b;
}
```

**R code**
```{r addition-r}
Rcpp::sourceCpp("Code/addition.cpp")
addition(2, 2)
```

### Rcpp advantages

**Thanks to `Rcpp::sourceCpp()`**

- Compile the C++ code
- Export the function to the R session
- Direct interchange of R objects (including S3, S4) between R and C++
- ... (many more, see `vignette("Rcpp-package")`)

**In an R package**

- `Rcpp.package.skeleton()` to generate a new Rcpp package (modifying `DESCRIPTION` and `NAMESPACE`)
- `Rcpp::compileAttributes()` scans the C++ files for `Rcpp::export` attributes and generates the code required to make the functions available in R.

## RcppGSL for fast random draws

### GSL and RcppGSL

\bcols
\bcol{0.1\textwidth}
\includegraphics[height=1.5cm]{figs/logo_GNU.png}
\ecol
\bcol{0.9\textwidth}

**GNU Scientific Library**

- Numerical library for C and C++ programmers
- Reliable random number generator algorithms
- Thoroughly tested and fast random number distributions
- Linear algebra (matrices and vectors)
- <https://www.gnu.org/software/gsl/>

\vspace{0.5cm}

**RcppGSL**

- Interface between R and GSL
- Using Rcpp to interface R and C
- <http://dirk.eddelbuettel.com/code/rcpp.gsl.html>

\ecol
\ecols

### GSL random number distributions

- GSL v2.6 includes **38 random number distributions** (see [GNU GSL](https://www.gnu.org/software/gsl/doc/html/randist.html))
- It's easy to implement additional random number distributions from the GSL base distributions (e.g. truncated normal distribution)
- For comparison, R API includes "only" 24 random number distributions (see [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Distribution-functions))
- Random draws are faster with GSL than with R (eg. `gsl_ran_gamma()` vs. `R::rgamma()`)

### RcppGSL example

**C++ code**
```{Rcpp RcppGSL-my_rnorm}
#include <Rcpp.h>
#include <gsl/gsl_rng.h>
#include <gsl/gsl_randist.h>

using namespace Rcpp;

// [[Rcpp::depends(RcppGSL)]]

// [[Rcpp::export]]
Rcpp::NumericVector my_rnorm(int nsamp, double mu,
                             double sigma) {
	gsl_rng *s = gsl_rng_alloc(gsl_rng_mt19937); // Random seed
	Rcpp::NumericVector beta(nsamp);
	for (int i = 0; i < nsamp; i++) {
		beta[i] = mu + gsl_ran_gaussian(s, sigma); // Random draw
	}
	return beta;
}
```

### RcppGSL example

**R code**
```{r r-my_rnorm, fig.show="asis", out.width="70%"}
library(Rcpp)
library(RcppGSL)
beta <- my_rnorm(100, 5, 2)
par(cex=2)
hist(beta)
```

## RcppArmadillo for high-performance linear algebra

### Armadillo and RcppArmadillo

\bcols
\bcol{0.1\textwidth}
\includegraphics[height=1.5cm]{figs/logo_Armadillo.png}  
\ecol
\bcol{0.9\textwidth}

**Armadillo**

- C++ library for linear algebra and scientific computing
- Provides high-level syntax and functionality: speed and ease of use
- Classes for vectors, matrices and cubes
- Matrix operations, matrix decomposition, linear model solver, etc.
- <http://arma.sourceforge.net/>

\vspace{0.5cm}

**RcppArmadillo**

- Interface between R and Armadillo
- Using Rcpp to interface R and C++
- <http://dirk.eddelbuettel.com/code/rcpp.armadillo.html>

\ecol
\ecols

### RcppArmadillo example

**C++ code**
```{Rcpp RcppArma-example}
#include <RcppArmadillo.h>

// [[Rcpp::depends(RcppArmadillo)]]

// [[Rcpp::export]]
Rcpp::List fastLm(const arma::mat& X, const arma::colvec& y) {
    int n = X.n_rows, k = X.n_cols;
        
    arma::colvec coef = arma::solve(X, y);    // fit model y ~ X
    arma::colvec res  = y - X*coef;           // residuals

    // std.errors of coefficients
    double s2 = std::inner_product(res.begin(), 
                                   res.end(), 
                                   res.begin(), 0.0)/(n - k);
                                                        
    arma::colvec std_err = arma::sqrt(s2 * 
      arma::diagvec(arma::pinv(arma::trans(X)*X)));

    return Rcpp::List::create(Rcpp::Named("coefficients") = coef,
                        Rcpp::Named("stderr")       = std_err,
                        Rcpp::Named("df.residual")  = n - k);
}
```

### RcppArmadillo example

**R code**
```{r r-RcppArma-example}
library(Rcpp)
library(RcppArmadillo)
# Trees data-set
y <- log(trees$Volume)
X <- cbind(1, log(trees$Girth))
# fastLm
mod <- fastLm(X, y)
mod$coef
```

### Licenses

- Licenses: GNU General Public License, Apache License 2.0 for Armadillo
- Free software licenses: we can use, modify and redistribute these softwares

<!-- ================================================================= -->
<!-- PART 2: Rcpp for Gibbs sampling -->
<!-- ================================================================= -->

# Rcpp for Gibbs sampling

## Gibbs sampling and Bayesian statistics

### Gibbs sampling

Gibbs sampling is commonly used for **statistical inference**.

$\Theta=(\theta_1,\ldots,\theta_n)$ from a joint distribution $p(\theta_1,\ldots,\theta_n)$.

1. Begin with some initials values $\Theta^{(i)}$.
2. Next sample $\Theta^{(i+1)}=(\theta_1^{(i+1)},\ldots,\theta_n^{(i+1)})$?     
We sample iteratively the parameters.     
Update $\theta_j^{(i+1)}$ according to $p(\theta_j^{(i+1)}|\theta_0^{(i+1)},\ldots,\theta_{j-1}^{(i+1)},\theta_{j+1}^{(i)},\ldots,\theta_{n}^{(i)})$
3. We repeat the above steps $k$ times (Markov chain Monte Carlo).

The samples approximate the joint distribution of all variables.

Can incorporate various algorithms (Metropolis-Hastings, slice sampling, adaptive rejection sampling, Hamiltonian Monte-Carlo) to implement one or more of the sampling steps.

### Gibbs sampling

- Involves several loops: MCMC step, parameters
- Random draws
- Matrix computations in case of conjugated priors in Bayesian statistics

Rcpp (C++), RcppGSL (random draws) and RcppArmadillo (matrix computations) are useful for efficient Gibbs sampling.

## Simple Gibb sampler

### R

**R code**
```{r Gibbs_R}
gibbs_r <- function(N, thin) {
	mat <- matrix(nrow = N, ncol = 2)
	x <- y <- 0
	
	for (i in 1:N) {
		for (j in 1:thin) {
			x <- rgamma(1, 3, y * y + 4) # Gamma(shape, rate) with R
			y <- rnorm(1, 1 / (x + 1), 1 / sqrt(2 * (x + 1)))
		}
		mat[i, ] <- c(x, y)
	}
	mat
}
```

### Rcpp

**C++ code with Rccp**
```{Rcpp Gibbs_Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericMatrix gibbs_rcpp(int N, int thin) {
	NumericMatrix mat(N, 2);
	double x = 0, y = 0;
	
	for(int i = 0; i < N; i++) {
		for(int j = 0; j < thin; j++) {
			x = R::rgamma(3, 1 / (y * y + 4)); // R::rgamma(shape, scale)
			y = R::rnorm((1 / (x + 1)), 1 / sqrt(2 * (x + 1)));
		}
		mat(i, 0) = x;
		mat(i, 1) = y;
	}
	
	return(mat);
}
```

### Rcpp $+$ RcppArmadillo $+$ RcppGSL 

**C++ code with RccpArmadillo and RcppGSL**
```{Rcpp Gibbs_Rcpp_Arma_GSL}
#include <RcppArmadillo.h>
#include <gsl/gsl_rng.h>
#include <gsl/gsl_randist.h>
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::depends(RcppGSL)]]

// [[Rcpp::export]]
arma::mat gibbs_rcpp_arma_gsl(int N, int thin) {
	gsl_rng *s = gsl_rng_alloc(gsl_rng_mt19937); // Create RNG seed
	arma::mat mat; mat.zeros(N, 2);
	double x = 0, y = 0;
	for(int i = 0; i < N; i++) {
		for(int j = 0; j < thin; j++) {
			x = gsl_ran_gamma(s, 3,
                     1 / (y * y + 4)); // Gamma(shape, scale)
			y = (1 / (x + 1)) + 
				gsl_ran_gaussian_ziggurat(s, 1 / sqrt(2 * (x + 1)));
		}
		mat(i, 0) = x;
		mat(i, 1) = y;
	}
	gsl_rng_free(s); // Free memory
	return(mat);
}
```

### Benchmark

```{r Benchmark}
# Libraries
library(rbenchmark)
# Benchmark
Benchmark <- benchmark(
	"R" = {gibbs_r(100, 10)},
	"rcpp" = {gibbs_rcpp(100, 10)},
	"rcpp_arma_gsl" = {gibbs_rcpp_arma_gsl(100, 10)},
	replications=30,
	columns = c("test", "elapsed", "relative")
)
Benchmark
```

## Linear regression example

<!-- ================================================================= -->
<!-- PART 3: jSDM R package -->
<!-- ================================================================= -->

# The `jSDM` R package

## Joint Species Distribution Models

### JSDM utility

\bcols
\bcol{0.6\textwidth}

- Fit species distribution models
- Accounting for species interaction

\includegraphics[width=\textwidth]{figs/allfrogs1.jpg}

\vspace{0.5cm}

- Can be used to explain/predict species range and produce species range map

\includegraphics[width=\textwidth]{figs/SDM.jpg}

\ecol
\bcol{0.4\textwidth}

\includegraphics[width=\textwidth]{figs/Litoria_ewingii.jpg}     

\ecol
\ecols

### Data to fit JSDM

- Species presence/absence on sites
- Environmental variables (climate, lancover) at each site

\vspace{0.5cm}

```{r jSDM_data, echo=FALSE}
df <- read.csv("tabs/jSDM_tab.csv")
df
```

## Model specification

### Statistical model

$\blue{Y}=\{\blue{y_{ij}}\}^{i=1,\ldots,nsite}_{j=1,\ldots,nsp}$, with:

$$\blue{y_{ij}}=\begin{cases}
    0 & \text{ if species $j$ is absent on site $i$}\\
    1 &  \text{ if species  $j$ is present on site $i$}.
    \end{cases}$$
    
We assume $\blue{y_{ij}} \sim \mathcal{B}ernouilli(\theta_{ij})$, with:

$$\text{probit}(\theta_{ij}) =\red{\alpha_i} + \red{\beta_{0j}} + \blue{X_i}\red{\beta_j} + \red{W_i\lambda_j}$$
$\red{\alpha_i}$: site random effects, with $\red{\alpha_i} \sim \mathcal{N}(0,\red{V_{\alpha}})$    
$\blue{X_i}$: known environmental variables on site $i$    
$\red{W_i}$: latent variables for site $i$     
$\red{\beta_j, \lambda_j}$: species fixed effects

Latent variables $\red{W_i}$: missing predictors $+$ main axes of covariation across taxa (see Warton et al. 2015 <doi: [10.1016/j.tree.2015.09.007](https://doi.org/10.1016/j.tree.2015.09.007)>).

### Complexity of the model

\bcols
\bcol{0.6\textwidth}

- Multi-dimensionality: parameters $\red{\alpha_i}$ for sites and $\red{\beta_j, \lambda_j}$ for species
- Non Gaussian process
- Latent-variables: $\red{W_i}$
- Mixed model with site random effects $\red{\alpha_i} \sim \mathcal{N}(0,\red{V_{\alpha}})$

\ecol
\bcol{0.4\textwidth}

\includegraphics[width=\textwidth]{figs/complexity.jpg} 

\ecol
\ecols

### jSDM R package

\includegraphics[width=\textwidth]{figs/jSDM-website.png}
\vspace{0.5cm}
<https://ecology.ghislainv.fr/jSDM>

## Comparison with `boral`/`JAGS`

