---
classoption: table, dvipsname, compress
output: 
  beamer_presentation:
    # keep_tex: true
    # toc: true
    slide_level: 3
    includes:
      in_header: header.tex
      before_body: doc_prefix.tex
      after_body: doc_suffix.tex
colorlinks: no
fontsize: 10pt
link-citations: yes
linkcolor: Black
urlcolor: Maroon
---


```{r options, include=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(echo=TRUE, cache=TRUE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```


# Introduction

## JSDMs

### Joint Species Distribution Models (JSDMs)

**Species Distribution Model (SDM), for one single species.**

- $y_i \sim Bernoulli(\theta_i)$, $y_i \in \{0,1\}$
- $i$: site
- $p(\theta_i) = X_i \beta$
- $X$: environmental variables
- $\beta$: species effects
  
**JSDM $=$ SDM for community of species.**

- $p(\theta_{ij}) = \alpha_i + X_i \beta_j + \Sigma_{ij}$
- $i$: site, $j$: species
- Site effect $\alpha_i$: mean site suitability
- Variance-covariance matrix $\Sigma_{ij}$: species co-occurrences

### Joint Species Distribution Models

JSDMs provide a convenient statistical framework to test **trait-environment** interactions.

$\beta_j$ can be expressed as a function of functional traits

- $p(\theta_{ij}) = \alpha_i + X_i \beta_j + \Sigma_{ij}$
- $p(\beta_j) = N(T_j \gamma, V_{\beta})$

JSDMs can help narrow the gap between **correlative** and **mechanistic** species distribution models.

`jSDM` R package (first chapter of Jeanne's PhD thesis), https://ecology.ghislainv.fr/jSDM/

## METRADICA's objectives

### Objectives of METRADICA (Task 3)

Using JSDMs:

- Test **trait-environmment** interactions for determining tree species distribution in French Guiana.
- Assess species vulnerability to climate change (through contraction of species range).
- Interpret species vulnerability to climate change in terms of functional traits.
- Derive maps of $\alpha$ and $\beta$ diversity for French Guiana.
- Identify refuge area for biodiversity under climate change (stable tree communities).

<!-- ============================================ -->
<!-- Methods -->
<!-- ============================================ -->

# Material and methods

## Datasets

### Datasets

\bcols
\bcol{0.5\textwidth}

Three types of data-sets:

- Species occurrences on sites
- Species trait database
- Environmental database

\ecol
\bcol{0.5\textwidth}
\centering \includegraphics[width=5cm]{figs/four-corner-model}
\ecol
\ecols

### Occurrences

```{r inventories, include=FALSE}
## Libraries
library(readr)
## Guyafor
f <- "/home/ghislain/Documents/Research_current/PROJECTS/METRADICA/Dataverse/Inventaires/202110_DonneesModeleMetradicaTout+Diam.csv"
df <- read_delim(f, delim=";")
nplots_1 <- length(unique(paste0(df$Project, df$idPlot)))
## Guyadiv
f <- "/home/ghislain/Documents/Research_current/PROJECTS/METRADICA/Dataverse/Inventaires/20211203_GuyaDiv.tab"
df <- read_tsv(f)
nplots_2 <- length(unique(paste0(df$Forest, df$idPlot)))
## Total
nplots <- nplots_1 + nplots_2
```

\bcols
\bcol{0.5\textwidth}

- Forest plot inventories coming from several networks combined together
- Networks: Guyafor, Gentry, Habitat, Guyadiv
- Presence-absence data and abundances
- `r nplots` forest plots
- About 1700 tree species, most of which are rare

\ecol
\bcol{0.5\textwidth}
\includegraphics[width=\textwidth]{figs/inventories}
\ecol
\ecols

### Traits

- Large "soft" trait (WD, LSA, tree max height, etc.) databases from previous CEBA projects.
- Five additional mechanistic traits from Metradica project:
  - leaf water potential at which cells lose turgor (Ptlp), minimum leaf conductance (gmin), leaf saturated water content (LSWC), vein density (VLA), stomatal density (SD).
  - 24 species, 672 trees, three sites with both hills and valleys spread on a precipitation gradient.

### Environment

- Topographic data (SRTM and LiDAR)
- Soil data
- Distance to human infrastructures (roads, villages)
- Climatic data (Chelsa) in the present and the future
- https://guyaclim.cirad.fr

## Study scales

### Scales: biogeography and micro-environment

**Local scale: microtopography $\times$ traits**

- Scale $=$ ~10km, resolution $=$ ~5m
- Explicative model: E $\times$ T
- Using MNT at 5m: hills (_"terra firme"_) and valleys

**Country scale (French Guiana)**

- Scale $=$ FG, resolution $=$ ~1km
- Explicative and predictive model
- Two models
  * Without traits
    - Predictive model
    - Present: distribution and co-occurrences of species
    - Future: range contraction in the future: (i) species vulnerability to climate change, (ii) change in species composition
  * With traits
    - Explicative model: E $\times$ T
    - Explaining species location (biogeography)

<!-- ============================================ -->
<!-- Perspectives -->
<!-- ============================================ -->

# Perspectives

## Model comparison

### Model comparison with forest dynamics models

**TROLL model**

- Tropical forest dynamics model
- Growth, mortality, recruitment through carbon allocation
- Species parameters are derived from traits
- Calibrated on some forests of French Guiana

**Model comparison**

- Species excluded from the community with TROLL under climate change.
- Do the same species experience a severe range contraction with JSDMs?

## Applications

### Applications

- Anticipating climate change effects on tropical forest in French Guiana
  - Massive tree mortality events and forest conversion to savannas?
  - Change in species composition?
- Identification of refuge areas for conservation $\Rightarrow$ systematic conservation planning.


<!-- ============================================ -->
<!-- EOF
<!-- ============================================ -->
