---
classoption: table, dvipsname, compress
output: 
  beamer_presentation:
    # keep_tex: true
    # toc: true
    slide_level: 3
    includes:
      in_header: header.tex
      before_body: doc_prefix.tex
      after_body: doc_suffix.tex
colorlinks: no
fontsize: 10pt
link-citations: yes
linkcolor: Black
urlcolor: Maroon
---

```{r options, include=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(echo=TRUE, cache=TRUE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```

# Introduction
<!-- ================================================================= -->
<!-- PART 1: Introduction -->
<!-- ================================================================= -->

## State of the art

### Available R packages for JSDMs

**Community of coexisting R/Python packages**

- **boral** (Warton and Hui)
- **HMSC** (Ovaskainen and Tikhonov)
- **gjam** (Clark and Gelfand)
- **BayesComm** (Golding)
- **s-jSDM** (Hartig and Pichler)
- \ldots

**Wilkinson, D. P.; Golding, N.; Guillera-Arroita, G.; Tingley, R.; McCarthy, M. A.; Peres-Neto, P.** 2018. A comparison of joint species distribution models for presence-absence data _Methods in Ecology and Evolution_, **10**:198-211. [doi: [10.1111/2041-210x.13106](https://doi.org/10.1111/2041-210x.13106)].

**Pichler, M.; Hartig F.** 2020. A new method for faster and more accurate inference of species associations from novel community data. _arXiv_ pre-print, https://arxiv.org/abs/2003.05331.

### Limitations

- Computational speed (boral, HMSC)
- Model specifications (BayesComm, s-jSDM)
	- eg. site random effects, functional traits, phylogenetic data
- Heterogenous model performance (HMSC, boral, gjam)

\centering \includegraphics[width=8cm]{figs/Norberg2019-EM.png}

Norberg et al. 2019

## Obectives

### Obectives of the jSDM R package

- Make our hands dirty to understand better the JSDM functioning
- Optimized code for fast MCMC computations
- User friendly: package, functions, articles, vignettes
- **A base for testing a large variety of models**:
  - occurrence and count data (Bernoulli/Binomial -- Poisson/Neg-Binomial)
	- probit/logit link function for occurrences
	- functional traits and phylogenetic data
	- species and site random/fixed effects
	- presence-only data
	
\vspace{0.25cm}

- Companion for the `hSDM` R package, hierarchical **one**-species distribution models (mixed models, imperfect detection, spatial autocorrelation) <https://ecology.ghislainv.fr/hSDM/>

<!-- ================================================================= -->
<!-- PART 2: jSDM R package -->
<!-- ================================================================= -->

# The `jSDM` R package

## Joint Species Distribution Models

### JSDM utility

\bcols
\bcol{0.6\textwidth}

- Fit species distribution models
- Accounting for species co-occurrences

\includegraphics[width=\textwidth]{figs/allfrogs1.jpg}

\vspace{0.5cm}

- Can be used to explain/predict species range and produce species range map

\includegraphics[width=\textwidth]{figs/SDM.jpg}

\ecol
\bcol{0.4\textwidth}

\includegraphics[width=\textwidth]{figs/Litoria_ewingii.jpg}     

\ecol
\ecols

### Data to fit JSDM

- Species presence/absence on sites
- Environmental variables (climate, lancover) at each site

\vspace{0.5cm}

```{r jSDM_data, echo=FALSE}
df <- read.csv("tabs/jSDM_tab.csv")
colnames <- c("Sites","Sp1","Sp2","...","Sp_nsp","X1","X2","...","X_nvar")
knitr::kable(df, col.names=colnames, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"), full_width=FALSE)
```

## Model specification

### Statistical model

$$\blue{y_{ij}}=\begin{cases}
    0 & \text{ if species $j$ is absent on site $i$}\\
    1 &  \text{ if species  $j$ is present on site $i$}.
    \end{cases}$$
    
We assume $\blue{y_{ij}} \sim \mathcal{B}ernoulli(\theta_{ij})$, with:

$$\text{probit}(\theta_{ij}) = \red{\alpha_i} + \red{\beta_{0j}} + \blue{X_i}\red{\beta_j} + \red{W_i\lambda_j}$$
$\red{\alpha_i}$: site random effects, with $\red{\alpha_i} \sim \mathcal{N}(0,\red{V_{\alpha}})$    
$\blue{X_i}$: known environmental variables on site $i$    
$\red{W_i}$: latent variables for site $i$
$\red{\beta_j, \lambda_j}$: species fixed effects

Latent variables $\red{W_i}$: missing predictors $+$ main axes of covariation across taxa (see Warton et al. 2015 <doi: [10.1016/j.tree.2015.09.007](https://doi.org/10.1016/j.tree.2015.09.007)>).

### Statistical model

The previous latent variable model (LVM):

$$\text{probit}(\theta_{ij}) = \red{\alpha_i} + \red{\beta_{0j}} + \blue{X_i}\red{\beta_j} + \red{W_i\lambda_j}$$

is equivalent to a multivariate probit regression (MPR):

$$\text{probit}(\theta_{ij}) = \red{\alpha_i} + \red{\beta_{0j}} + \blue{X_i}\red{\beta_j} + \red{u_{ij}}$$

with $\red{u_{ij}} \sim \mathcal{N}(0, \Sigma)$ (where $\Sigma$ is the variance-covariance matrix) and with the constraint that $\Sigma = \Lambda \Lambda^{\prime}$ (where $\Lambda$ is the full matrix of factor loadings, with the $\lambda_j$ as its columns).

### Complexity of the model

\bcols
\bcol{0.6\textwidth}

- Multi-dimensionality: parameters $\red{\alpha_i}$ for sites and $\red{\beta_j, \lambda_j}$ for species
- Non Gaussian process
- Latent-variables $\red{W_i}$
- Mixed model with site random effects $\red{\alpha_i} \sim \mathcal{N}(0,\red{V_{\alpha}})$

\ecol
\bcol{0.4\textwidth}

\includegraphics[width=\textwidth]{figs/complexity.jpg} 

\ecol
\ecols

### jSDM R package

\includegraphics[width=\textwidth]{figs/jSDM-website.png}

- <https://ecology.ghislainv.fr/jSDM>
- Made with Rcpp* packages

## Rcpp* packages

### Rcpp R package

- **Rcpp** is an R package to extend R with C++ code
- Main advantage: C++ is fast, it accelerates R (see next sections)
- Written by **Dirk EDDELBUETTEL** and **Romain FRANCOIS**
- <http://www.rcpp.org/>

### Simple Rcpp example

**C++ code** (in file `Code/addition.cpp`)
```{Rcpp Rcpp-addition, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
int addition(int a, int b) {
  return a + b;
}
```

**R code**
```{r addition-r}
Rcpp::sourceCpp("Code/addition.cpp")
addition(2, 2)
```

### Rcpp advantages

**Thanks to `Rcpp::sourceCpp()`**

- Compile the C++ code
- Export the function to the R session
- Direct interchange of R objects (including S3, S4) between R and C++
- ... (many more, see `vignette("Rcpp-package")`)

**In an R package**

- `Rcpp.package.skeleton()` to generate a new Rcpp package (modifying `DESCRIPTION` and `NAMESPACE`)
- `Rcpp::compileAttributes()` scans the C++ files for `Rcpp::export` attributes and generates the code required to make the functions available in R.

### GSL and RcppGSL for fast random draws

\bcols
\bcol{0.1\textwidth}
\includegraphics[height=1.5cm]{figs/logo_GNU.png}
\ecol
\bcol{0.9\textwidth}

**GNU Scientific Library**

- Numerical library for C and C++ programmers
- Reliable random number generator algorithms
- Thoroughly tested and fast random number distributions
- Linear algebra (matrices and vectors)
- <https://www.gnu.org/software/gsl/>

\vspace{0.5cm}

**RcppGSL**

- Interface between R and GSL
- Using Rcpp to interface R and C
- <http://dirk.eddelbuettel.com/code/rcpp.gsl.html>

\ecol
\ecols

### GSL random number distributions

- GSL v2.6 includes **38 random number distributions** (see [GNU GSL](https://www.gnu.org/software/gsl/doc/html/randist.html))
- It's easy to implement additional random number distributions from the GSL base distributions (e.g. truncated normal distribution)
- For comparison, R API includes "only" 24 random number distributions (see [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Distribution-functions))
- Random draws are faster with GSL than with R (eg. `gsl_ran_gamma()` vs. `R::rgamma()`)

### Armadillo and RcppArmadillo for high-performance linear algebra

\bcols
\bcol{0.1\textwidth}
\includegraphics[height=1.5cm]{figs/logo_Armadillo.png}  
\ecol
\bcol{0.9\textwidth}

**Armadillo**

- C++ library for linear algebra and scientific computing
- Provides high-level syntax and functionality: speed and ease of use
- Classes for vectors, matrices and cubes
- Matrix operations, matrix decomposition, linear model solver, etc.
- <http://arma.sourceforge.net/>

\vspace{0.5cm}

**RcppArmadillo**

- Interface between R and Armadillo
- Using Rcpp to interface R and C++
- <http://dirk.eddelbuettel.com/code/rcpp.armadillo.html>

\ecol
\ecols

### GSL and Armadillo licenses

- Licenses: GNU General Public License, Apache License 2.0 for Armadillo
- Free software licenses: we can use, modify and redistribute these softwares

<!-- ================================================================= -->
<!-- PART 3: -->
<!-- ================================================================= -->

# Comparison with `boral`/`JAGS`

## `boral` R package

### `boral` R package

- R package interfacing R with JAGS for fitting Joint Species Distribution Models
- JAGS is Just Another Gibbs Sampler: <http://mcmc-jags.sourceforge.net/>
- Approach used in Warton et al. 2015:    
<doi: [10.1016/j.tree.2015.09.007](https://doi.org/10.1016/j.tree.2015.09.007)>
- boral by Francis K.C. Hui and JAGS by Martyn Plummer

## Data-sets

### Data-sets

```{r data-sets, echo=FALSE}
library(dplyr)
datasets <- read.table("tabs/data-sets.csv", header=TRUE, sep=",")
datasets <- datasets %>%
	dplyr::mutate(npar=nsite*(1+nlat)+1+nsp*(1+nX+nlat)-1, nobs=nsite*nsp) %>%
	dplyr::select(dataset,nsite,nsp,nobs,nX,nlat,npar,nmcmc)
knitr::kable(datasets, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"), full_width=FALSE)
```

\vspace{0.5cm}

\begin{center}
\begin{tabular}{cccc}
\includegraphics[width=2cm]{figs/Mosquitos_.jpg} &
\includegraphics[width=2cm]{figs/Eucalyptus_.jpg} &
\includegraphics[width=2cm]{figs/Frogs_.jpg} &
\includegraphics[width=2cm]{figs/Fungi_.jpg} \\
\textbf{Mosquitos} & \textbf{Eucalyptus} & \textbf{Frogs} & \textbf{Fungi}
\end{tabular}
\end{center}

## Results

### Comparison results

**Compilation time** (in minutes)
```{r time, echo=FALSE}
load("Code/jSDM_boral_sim.RData")
load("Code/jSDM_boral_mosquito.RData")
load("Code/jSDM_boral_eucalypts.RData")
load("Code/jSDM_boral_frogs.RData")
load("Code/boral_fungi.RData")
load("Code/jSDM_fungi.RData")

result <- data.frame(matrix(NA,2,5), row.names=c("boral","jSDM "))
colnames(result) <- c("Simulated","Mosquitos","Eucalyptus","Frogs","Fungi")
result[1,] <- c(T_boral_sim, T_boral_Mosquito, T_boral_Eucalypts, T_boral_Frogs, T_boral_Fungi)
result[2,] <- c(T_jSDM_block_sim, T_jSDM_block_Mosquito, T_jSDM_block_Eucalypts, T_jSDM_block_Frogs, T_jSDM_block_Fungi)
ratio_speed <- round(mean(as.numeric(result[1,]/result[2,])))
min_ratio <- round(min(as.numeric(result[1,]/result[2,])))
max_ratio <- round(max(as.numeric(result[1,]/result[2,])))
knitr::kable(result, digits =1, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"), font_size=8, full_width=FALSE)
```

jSDM is **`r min_ratio` to `r max_ratio`** times faster than boral/jags.

**Root-mean-square error**    

Computed for $\text{probit}(\theta_{ij})$ with the simulated data-set.

```{r RMSE, echo=FALSE}
library(knitr)
result <- data.frame(matrix(NA,1,2),row.names=c("RMSE"))
colnames(result) <- c("boral","jSDM")
result$boral <- RMSE_boral_sim
result$jSDM <- RMSE_jSDM_block_sim
knitr::kable(result, digits=1, booktabs=TRUE) %>%
	kableExtra::kable_styling(latex_options=c("HOLD_position","striped"), font_size=8, full_width=FALSE)
```

**Deviance**

```{r deviance, echo=FALSE}
library(knitr)
result <- data.frame(matrix(NA,2,5),row.names=c("boral", "jSDM"))
colnames(result) <- c("Simulated","Mosquitos","Eucalyptus","Frogs","Fungi")
result[1,] <- c(Deviance_boral_sim, Deviance_boral_Mosquito, Deviance_boral_Eucalypts, Deviance_boral_Frogs, Deviance_boral_Fungi)
result[2,] <- c( mean(mod_jSDM_block_sim$mcmc.Deviance),  mean(mod_jSDM_block_Mosquito$mcmc.Deviance),  mean(mod_jSDM_block_Eucalypts$mcmc.Deviance),  mean(mod_jSDM_block_Frogs$mcmc.Deviance), mean(mod_jSDM_block_Fungi$mcmc.Deviance))
knitr::kable(result, digits=0, booktabs=TRUE) %>%
	kableExtra::kable_styling(latex_options=c("HOLD_position","striped"), font_size=8, full_width=FALSE)
```

### Conclusion

- Small data-sets **and** simple models: R, *BUGS, JAGS, Stan, INLA, MCMCglmm
- Large data-sets **or** complex hierarchical models: R $+$ Rcpp $+$ RcppGSL $+$ RcppArmadillo

\vspace{0.5cm}

- With Rcpp* packages, the Gibbs sampler can typically be written in about half a day
- Code is reusable and easily packageable
- Tools with incomparable efficiency for statisticians

# Perspectives

## Additional functionalities

### Additional functionalities

- Count data (Poisson/Negative-Binomial)
- Logit link function for occurrences
- Functional traits and phylogenetic data
- Species and site random/fixed effects
- Presence-only data
- Spatial autocorrelation for $\alpha_i$ and $W_i$

## SDM vs. JSDM

### SDM vs. JSDM

See notebook
