---
classoption: table, dvipsname, compress
output: 
  beamer_presentation:
    # keep_tex: true
    # toc: true
    slide_level: 3
    includes:
      in_header: header.tex
      before_body: doc_prefix.tex
      after_body: doc_suffix.tex
colorlinks: no
fontsize: 10pt
link-citations: yes
linkcolor: Black
---

```{r options, echo=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(echo=FALSE, cache=FALSE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```

# Introduction

## Long term projections

### Long term projections

\bcols
\bcol{0.5\textwidth}

- Tropical forests shelter most of the terrestrial biodiversity and carbon stocks
- They are currently being deforested at rates close to **1%/yr**

\ecol
\bcol{0.5\textwidth}

\centering \includegraphics[width=5cm]{figs/fcc13_COD}
**2005-2015** deforestation in Democratic Republic of the Congo

\ecol
\ecols

\vspace{0.5cm}

\centering What happens when you project annual deforestation on the medium to long term (2050-2100)?

$\rightarrow$ Long term projections

## Spatial projections

### Spatial projections

\bcols
\bcol{0.5\textwidth}

- Not all forests are equally threathened
- And biodiversity and forest carbon stocks vary spatially

\ecol
\bcol{0.5\textwidth}

\centering \includegraphics[width=5cm]{figs/agb_COD}
Aboveground biomass in Democratic Republic of the Congo

\ecol
\ecols

\vspace{0.5cm}

\centering What are the consequences of long term deforestation for biodiversity and CO2 emissions?

$\rightarrow$ Long term **spatial** projections

## Objectives

### Objectives

- Deriving high-resolution maps of the spatial probability of deforestation
- Projecting forest cover change until 2050 under a business-as-usual scenario

# Data

## Historical deforestation

### Historical deforestation

- Wall-to-wall map of **tropical moist forest cover** change at 30 m resolution from 1990 to 2018
- Using the 37-years full Landsat satellite archive and Google Earth Engine
- Time-series analysis at the pixel scale using a complex decision tree based on expert knowledge

\centering \includegraphics[width=10cm]{figs/Christelle_Africa}

### Historical deforestation

\centering \includegraphics[width=6.5cm]{figs/Deforestation_Patterns}

### Historical deforestation

- **Vancutsem Ch., F. Achard , J.-F. Pekel , G. Vieilledent, S. Carboni , D. Simonetti , J. Gallego.** Long-term monitoring of the tropical moist forests dynamics reveals unprecedented deforestation rates. Submitted to _Nature Communications_.

- Response variale: deforestation on 2005-2015

## Explicative variables

### Explicative variables

- Variable types: **landscape**, **accessibility**, **protection status**

```{r variables}
tab1 <- read.table("tables/variables.txt",sep=",",header=TRUE)
names(tab1) <- c("Product","Source","Variable derived","Unit","Resolution (m)")
# Kable
knitr::kable(tab1, linesep="", booktabs=TRUE) %>%
  kable_styling(latex_options = c("striped"), font_size=7) %>%
  column_spec(1, width = "2cm") %>%
  column_spec(3, width = "2cm")
```

$(1)$ Vancutsem et al., $(2)$ <http://srtm.csi.cgiar.org>, $(3)$ <http://www.geofabrik.de>, $(4)$ <http://protectedplanet.net> 

# Modelling

## Statistical model

### Statistical model

- $Y_{ij} \in \{0, 1\} \sim \mathcal{B}ernoulli(\theta_{ij})$
- $\logit(\theta_{ij}) = f(\text{spatial factors}_i) + \rho_j$
- Autocorrelated spatial random effects $\rho_j$ (10 km) to account for **unmeasured** or **unmeasurable** factors: population density, soil type, geographical barriers, law enforcement locally
- Hierarchical Bayesian framework

\centering \includegraphics[height=3cm]{figs/iCAR}\hspace{1cm}
\centering \includegraphics[height=3cm]{figs/Bayes}

### Statistical model

- One model per country
- 40,000 sample points (balanced sampling deforested/non-deforested areas)
- Variable selection based on expert knowledge regarding effect sign

## Software

### Software

\centering \includegraphics[width=\textwidth]{figs/forestatrisk_pkg}

- `forestatrisk` Python package: <https://github.com/ghislainv/forestatrisk>
- Rasters processed by chuncks: high resolution (30 m, large spatial scale)
- Fast, without memory issues
- Parallel computation: one node per country

# Results




