---
classoption: table, dvipsname, compress
output: 
  beamer_presentation:
    # keep_tex: true
    # toc: true
    slide_level: 3
    includes:
      in_header: header.tex
      before_body: doc_prefix.tex
      after_body: doc_suffix.tex
colorlinks: no
fontsize: 10pt
link-citations: yes
linkcolor: Black
---

```{r options, echo=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(echo=FALSE, cache=FALSE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```

# Introduction

## Context

### Long term projections

\bcols
\bcol{0.5\textwidth}

- Tropical forests shelter most of the terrestrial biodiversity and carbon stocks
- They are currently being deforested at rates close to **1%/yr**

\ecol
\bcol{0.5\textwidth}

\centering \includegraphics[width=5cm]{figs/fcc13_COD}
**2005-2015** deforestation in Democratic Republic of the Congo

\ecol
\ecols

\vspace{0.5cm}

\centering What happens when you project annual deforestation on the medium or long term (2050-2100)?

### Spatial projections

\bcols
\bcol{0.5\textwidth}

- Not all forests are equally threathened
- And biodiversity and forest carbon stocks vary spatially

\ecol
\bcol{0.5\textwidth}

\centering \includegraphics[width=5cm]{figs/agb_COD}
Aboveground biomass in Democratic Republic of the Congo

\ecol
\ecols

\vspace{0.5cm}

\centering What are the consequences of long term deforestation for biodiversity and CO2 emissions?

## Objectives

### Objectives

- Modelling the deforestation process spatially
- Deriving high-resolution maps of the spatial probability of deforestation
- Projecting forest cover change until 2050 under a business-as-usual scenario
- At the pantropical scale

\centering \includegraphics[width=8cm]{figs/Moist_Tropical_Forest}

# Data

## Historical deforestation

### Historical deforestation

- Wall-to-wall map of **tropical moist forest cover change** at 30 m resolution from 1990 to 2018
- Using the 37-years full Landsat satellite archive and Google Earth Engine
- Time-series analysis at the pixel scale using a complex decision tree based on expert knowledge

\centering \includegraphics[width=10cm]{figs/Christelle_Africa}

### Historical deforestation

\centering \includegraphics[width=6.5cm]{figs/Deforestation_Patterns}

### Historical deforestation

- **Vancutsem Ch., F. Achard , J.-F. Pekel , G. Vieilledent, S. Carboni , D. Simonetti , J. Gallego.** Long-term monitoring of the tropical moist forests dynamics reveals unprecedented deforestation rates. Submitted to _Nature Communications_.

- Hansen et al. 2013: underestimated deforestation rates in Africa (small scale mosaic deforestation)

- Response variable: deforestation on 2005-2015

## Explicative variables

### Explicative variables

- Variable types: **landscape**, **accessibility**, **protection status**

```{r variables}
tab1 <- read.table("tables/variables.txt",sep=",",header=TRUE)
names(tab1) <- c("Product","Source","Variable derived","Unit","Resolution (m)")
# Kable
knitr::kable(tab1, linesep="", booktabs=TRUE) %>%
  kable_styling(latex_options = c("striped"), font_size=7) %>%
  column_spec(1, width = "2cm") %>%
  column_spec(3, width = "2cm")
```

$(1)$ Vancutsem et al., $(2)$ <http://srtm.csi.cgiar.org>, $(3)$ <http://www.geofabrik.de>, $(4)$ <http://protectedplanet.net> 

# Modelling

## Statistical model

### Statistical model

- $\textcolor{blue}{Y_{ij}} \in \{0, 1\} \sim \mathcal{B}ernoulli(\theta_{ij})$
- $\logit(\theta_{ij}) = \textcolor{blue}{X_i}\textcolor{red}{\beta} + \textcolor{red}{\rho_j}$

- Autocorrelated spatial random effects $\textcolor{red}{\rho_j}$ (10 km) to account for **unmeasured** or **unmeasurable** factors: population density, soil type, geographical barriers, law enforcement locally
- Structure spatially the **residual variability** that is not explained by the model's variables
- Hierarchical Bayesian framework

\centering \includegraphics[height=3cm]{figs/iCAR}\hspace{1cm}
\centering \includegraphics[height=3cm]{figs/Bayes}

### Statistical model

- One model per country
- 40,000 sample points (balanced sampling deforested/non-deforested areas)
- Variable selection (statistical significance $+$ process coherence)

## Software

### Software

\centering \includegraphics[width=\textwidth]{figs/forestatrisk_pkg}

- `forestatrisk` Python package: <https://github.com/ghislainv/forestatrisk>
- Rasters processed by chuncks: high resolution (30 m, large spatial scale)
- Fast, without memory issues
- Parallel computation: one node per country

# Results

## Parameters

### Parameters

```{r parameters, size="\\tiny"}
path <- "tables/summary_hSDM_COD.txt"
conn <- file(path, open="r")
#writeLines(readLines(conn))
close(conn)
```

Parameter values: $\textcolor{red}{\beta}$ and variance $\textcolor{red}{V_{\rho}}$ of the spatial random effects.

\centering \includegraphics[width=\textwidth]{figs/variable_effects}

- Set of parameters for each country.
- Each effect can be easily interpreted.
- Effects can be compared between countries (efficiency of the protected areas, effect of road infrastructures).

### Spatial random effects

\centering \includegraphics[width=7cm]{figs/rho_orig_COD}

Then interpolated at 1km.

## Spatial probability

### Spatial probability of deforestation

\centering \includegraphics[width=6cm]{figs/prob_COD}

### Future forest cover

\centering \includegraphics[width=6cm]{figs/fcc_35yr_COD}

Projected forest cover change in **2015-2050** under a business-as-usual scenario.    
BAU: historical deforestation (ha/yr) observed on **2005-2015**.

### African continent

\centering \includegraphics[width=10cm]{figs/prob_Africa}

Spatial probability of deforestation.

### African continent

\centering \includegraphics[width=10cm]{figs/fcc_35yr_Africa.png}

Forest cover change in **2015-2050**, BAU scenario **2005-2015**.    

### African continent

\centering \includegraphics[width=6cm]{figs/fcc_35yr_Africa.png}

- **No more moist forests in 2050**: West-African countries except Liberia, East-African countries including Madagascar
- **Remaining forest block**: Congo, Gabon, Equatorial-Guinea, Cameroon
- **Highly fragmented forest**: Democratic Republic of the Congo
- **Two blocks of forest on both sides of the Congo River**

# Conclusion

## Projections

### Projections

- **Fake** and **unrealistic** scenario?
  - Simple BAU scenario **driven by observations** (deforestation on 2005-2015)
  - **Optimistic scenario**: no effect of the demographic growth in Africa in the model

\centering \includegraphics[width=6cm]{figs/UN-regional-pop-projections.png}
  
### Projections

- **Projections taken seriously** by the European Commission
- On 23 July 2019, the European Commission presented a **set of actions** to protect and restore the world's forests. [EC Communication](https://bit.ly/2JTAYRh).

\centering \includegraphics[width=7cm]{figs/EU}
  
## Perspectives

### Perspectives

- Maps can be used to identify **high-priority conservation areas** (biodiversity, carbon, etc.)
- **Additional scenario** taking into account the demographic growth
- Extending the work to the **entire tropics**
- **Spreading the word** to (hopefully) have impact on policies

\centering \includegraphics[width=7cm]{figs/tarzan.jpg}
